Automatically Splitting a Staged Lambda Calculus

### 2015-07-10 paper/splitting.tex
added equivalence figure

@@

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
They are simply recursive on the structure of the term, 
implying that the splitting algorithm runs in linear time.

Before discussing the correctness of these translations, 
we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting judgement defines a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
This output is a multistage program in its own right.
It is evaluated by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
In the base case, where $e$ is already a residual of the form $\exv q$, we know
$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
$p = \tup{}$ and $r = q$.
From here, we can directly derive $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
To help this case, we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.

This concludes the proof of program correctness, modulo \ref{lem:twoCorrect}.
However in order to prove that lemma, 
we will need to state analogous version for steps at \bbonem,
since the various kinds multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

Thus instead of a pipeline relationship, $i$ and $q$ bear more of a parallel relationship.
That is, they embody the complementary portions of $v$ that are relevant to each stage.
For this reason we call $i$ the \emph{immediate value} and $q$ the residual.

@@

\input{figures/splitting-term}

>>>>

\TODO add related work

@@

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
Since the rules are simply recursive on the structure of the term, 
the splitting algorithm runs in linear time on the size of the input program.
Splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting judgement defines a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
This output is evaluated in two passes by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

\TODO ADD EQUIVALENCE FIGURE

We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
In the base case, where $e$ is already a residual of the form $\exv q$, we know
$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, so by uniqueness of splitting, 
$p = \tup{}$ and $r = q$.
From here, we can directly derive $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.

\TODO defin the gap, then close it with the lemma
To help this case, we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.

From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.

The above proof of \ref{thm:allCorrect} invokes \ref{lem:twoCorrect},
In order to prove that lemma, 
we will need to state analogous version for steps at \bbonem,
since the various kinds multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

Thus $i$ and $q$ embody the complementary portions of $v$ that are relevant to each stage.
For this reason we call $i$ the \emph{immediate value} and $q$ the residual.

@@

\input{figures/splitting-term}
\input{figures/splitting-misc}
